Multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-08-10 paper/mcd2.tex
edits on multilevel FE CD constructs

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bF}{\mathbf{F}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bzero}{\bm{0}}

@@

\newcommand{\mR}{R^{\bm{\oplus}}}
\newcommand{\iR}{R^{\bullet}}

@@

We will approximately solve VI \eqref{eq:vi} for certain nonlinear obstacle problems, over two-dimensional computational domains $\Omega \subset \RR^2$ and the Banach space $\mathcal{V}=W_0^{1,p}(\Omega)$, using piecewise-linear finite element (FE) spaces over triangular meshes.  The targeted mesh will be the finest level in a hierarchy of meshes constructed by refinement of a coarsest mesh.

The mesh hierarchy and FE spaces are constructed in a standard manner \cite{Elmanetal2014}.  Suppose $\Omega$ is polygonal and triangulated by $\mathcal{T}^0$, the \emph{coarsest mesh}, a finite set of non-overlapping triangles with union $\overline{\Omega}$ and characteristic size (cell diameter) $h_0>0$.  Starting with $j=1$, let $\mathcal{T}^j$ be the uniform refinement of $\mathcal{T}^{j-1}$ by edge bisection, thus each triangle becomes four similar triangles in $\mathcal{T}^j$ with characteristic size $h_j = 2^{-j} h_0$.  Refinement proceeds to level $J\ge 0$, called the \emph{depth} of the hierarchy, giving $\mathcal{T}^J$ as the \emph{finest mesh}.  (In the single-level $J=0$ case all multilevel efficiency will be lost.)  Let $m_j$ be the number of interior nodes in $\mathcal{T}^j$, the \emph{degrees of freedom} at the $j$th level, and denote $h=h_J=2^{-J} h_0$ and $m=m_J$ for the finest level.

Let $\mathcal{V}^j$ be the continuous, piecewise-linear FE space $P_1$ over $\mathcal{T}^j$ with zero boundary values.  Note $\mathcal{V}^j$ is a subspace of $\mathcal{V}=W_0^{1,p}(\Omega)$ \cite{Elmanetal2014}, and observe the following nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}.  \label{eq:fe:nestedspaces}
\end{equation}

Let $\tilde{\mathcal{V}}^j$ denote the set of functions on the nodes of $\mathcal{T}^j$ which take their values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  Our purpose for such extended-value functions is to define ``box constraints'' \cite{BensonMunson2006} as follows.  At the finest-level, suppose $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad \overline{\gamma}^J > -\infty, \qquad \text{and} \qquad \underline{\gamma}^J|_{\partial\Omega} \le 0 \le \overline{\gamma}^J|_{\partial\Omega}. \label{eq:fe:boxconstraintrequirements}
\end{equation}
We call $\underline{\gamma}^J$ the \emph{lower obstacle} and $\overline{\gamma}^J$ the \emph{upper obstacle}.  Where $\underline{\gamma}^J=-\infty$ there is no lower obstacle and where $\overline{\gamma}^J=+\infty$ there is no upper obstacle.  Now define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,\underline{\gamma}^J \le v \le \overline{\gamma}^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions.  Note that admissible functions from $\mathcal{K}^J$ have zero boundary values and that elements of $\mathcal{K}^J$ never, themselves, take on values $\pm\infty$.

The FE method herein apparently extends to $Q_1$ spaces over quadrilateral meshes \cite{Elmanetal2014}.  However, for $k>1$ the $P_k$ space seems not to be suitable because the monotone restriction operator, \eqref{eq:fe:monotonerestriction} below, depends on element-wise monotonicity.  That is, for $P_1$ or $Q_1$ if the nodal values of an FE function (non-strictly) increase then the FE function also increases.  As shown by Figure 1.7 of \cite{Elmanetal2014}, for example, such monotonicity is not true for $P_2$.

One might assume that the nonlinear functional $f$ in \eqref{eq:vi} is defined over $\mathcal{K}^J$, i.e.~$f:\mathcal{K}^J \to \mathcal{V}'$.  However, especially for nontrivial nonlinear functionals, the finest-level, discretized FE functional $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ is likely to only approximate $f$ because of quadrature or other implementation details.  The functional $\ell$ in \eqref{eq:vi} is defined and continuous over $\mathcal{V}^J$, thus $\ell\in (\mathcal{V}^J)'$, and we assume for simplicity that it can be computed exactly, but we denote $\ell^J \in (\mathcal{V}^J)'$ to indicate that it must also be given an FE representation.

With these understandings we seek the finest-level FE solution $u^J \in \mathcal{K}^J$ of this finite-dimensional, nonlinear VI
\begin{equation}
\ip{f^J(u^J)}{v-u^J} \ge \ip{\ell^J}{v-u^J} \qquad \text{for all } v\in \cK^J. \label{eq:fe:vi}
\end{equation}
As for \eqref{eq:vi}, if $f^J$ is continuous and $p$-coercive on $\mathcal{K}^J$, in the sense that \eqref{eq:pcoercive} holds for all $u,v \in \mathcal{K}^J$, or $f^J$ is continuous, coercive and strictly monotone, then \eqref{eq:fe:vi} is well-posed \cite{KinderlehrerStampacchia1980}.

While \eqref{eq:fe:vi} approximates \eqref{eq:vi}, it may not technically be conforming for \eqref{eq:vi}.  That is, even if quadrature and other implementation details are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require $\mathcal{K}^J \subset \mathcal{K}$.  Only the fine-level FE obstacle $\gamma^J$ needs to be defined, while $\mathcal{K}$ remains abstract.  In fact, if a continuum obstacle exists then admissibility with respect to its FE interpolant $\gamma^J$ might not imply continuum admissiblity.  These caveats apply in geophysical situations where the obstacle is given by meshed elevation data, the continuum limit of which is undefined \cite{Bueler2016}.

The multilevel algorithm in the next section, designed to solve \eqref{eq:fe:vi}, requires transfer operators between meshes.  We use four such operators, three of which are standard, linear maps used for multigrid methods \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.

\begin{table}
\begin{tabular}{lll}
\emph{name}  & \emph{action}  & \emph{linear?} \\ \hline
canonical prolongation        & $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ & \,\checkmark \\
canonical (dual) restriction  & $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ & \,\checkmark \\
(nodal) injection restriction & $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$ & \,\checkmark \\
monotone restriction          & $\mR:\tilde{\mathcal{V}}^j\to\tilde{\mathcal{V}}^{j-1}$ &
\end{tabular}

\medskip
\caption{Transfer operators used in this paper.  FE vector space $\mathcal{V}^j$ has zero boundary values while $\tilde{\mathcal{V}}^j$ has arbitrary boundary values. $\mR$ is a nonlinear map.}
\label{tab:transfers}
\end{table}

Canonical prolongation $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ acts as the identity on coarser-mesh functions in $\mathcal{V}^{j-1}$; it is the linear vector space injection $\mathcal{V}^{j-1} \hookrightarrow \mathcal{V}^j$.  Canonical (dual) restriction $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ also acts as an identity in the sense that if $\sigma \in (\mathcal{V}^j)'$ then $(R\sigma)[z] = \sigma[z]$ for all $z \in \mathcal{V}^{j-1} \subset \mathcal{V}^j$.  While these canonical transfers do not lose information, they typically involve a change of FE representation.  For example, the practical formula for canonical restriction, derived by writing coarser-mesh hat functions in terms of finer-mesh ones, is a ``full-weighting'' average \cite{Trottenbergetal2001}.  It is the transpose of canonical prolongation.

To define injection $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$, observe that for any given values at the (interior) nodes of $\mathcal{T}^j$ there is a unique function in $\mathcal{V}^j$ with those values.  Also observe that, by our refinement construction, the nodes of $\mathcal{T}^{j-1}$ are also nodes of $\mathcal{T}^j$.  For $z\in\mathcal{V}^j$ we define the injection $\iR z$ as the element of $\mathcal{V}^{j-1}$ which has the same nodal values as $z$ at the nodes of $\mathcal{T}^{j-1}$.  Note that injection preserves ordering, that is, if $w \le z$ then $\iR w \le \iR z$.

Finally, monotone restriction $\mR:\tilde{\mathcal{V}}^j\to\tilde{\mathcal{V}}^{j-1}$ is defined as the nonlinear map which maximizes nodal values over the supports of coarse-mesh hat functions.  To make the definition precise, first denote the \emph{nodes} of triangulation $\mathcal{T}^j$ by $x_q^j$ for $q=0,\dots,\tilde{m}_j-1$, and then the corresponding hat functions \cite{Elmanetal2014} by $\psi_q^j$.  (Boundary nodes are included here.)  Monotone restriction of $z\in\tilde{\mathcal{V}}^j$ is
\begin{equation}
  \mR z = \sum_{p=1}^{\tilde m_{j-1}} \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}. \label{eq:fe:monotonerestriction}
\end{equation}
Note that if $z\in\tilde{\mathcal{V}}^j$ then $z \le \mR z$.

@@

To make this precise, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the \emph{defect constraint} \cite{GraeserKornhuber2009}
\begin{equation}
\chi^J = \gamma^J - w^J,  \label{eq:fe:defectconstraintfcn}
\end{equation}
a nonpositive function in $\tilde{\mathcal{V}}^J$.  The motivation here is that a corrected iterate is admissible, $w^J + z^J \ge \gamma^J$, if and only if the perturbation $z^J$ exceeds the defect constraint, $z^J \ge \chi^J$.  Next, for $j=J,J-1,\dots,1$ we apply monotone restriction $\mR$ to define the \emph{level defect constraints},
\begin{equation}
\chi^{j-1} = \mR \chi^j. \label{eq:fe:chilevels}
\end{equation}
so that $\chi^j \le \chi^{j-1}$.  A $J=4$ example is shown in Figure \ref{fig:chiphilevels}.

@@

\begin{figure}[h]
\begin{center}
\input{tikz/nmcdvcycle.tex}
\end{center}
\caption{An NMCD V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j$ expand into larger constraint subsets $\mathcal{U}^j$.}
\label{fig:nmcdvcycle}
\end{figure}

@@

\begin{pseudofloat}[H]
\begin{pseudo} \label{ps:nmcd-vcycle}
\pr{nmcd-vcycle}(\ell^J,\gamma^J;w^J)\text{:} \\+
    $\chi^J = \gamma^J - w^J$ \\
    $g^J = w^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \mR \chi^j$ \\
      $\phi^j = \chi^j - P\chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\phi^j,g^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(g^{j-1}) + R \left(\ell^j - f^j(g^j+y^j)\right)$ \\-
    $y^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\chi^0,g^0;y^0)$ \ct{solving of \eqref{eq:fe:downvi} in $\mathcal{D}^0=\mathcal{U}^0$} \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\chi^j,g^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\caption{Nonlinear multilevel constraint decomposition V-cycle for the finest-level VI problem \eqref{eq:fe:vi}, as an in-place method on $w^J$.  $f^j$ denotes a discretization of $f$ in problem \eqref{eq:vi}.}
\label{alg:nmcd}
\end{pseudofloat}

>>>>

\newcommand{\bb}{\mathbf{b}}
\newcommand{\be}{\mathbf{e}}
\newcommand{\bbf}{\mathbf{f}}
\newcommand{\bg}{\mathbf{g}}
\newcommand{\bn}{\mathbf{n}}
\newcommand{\br}{\mathbf{r}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}
\newcommand{\bw}{\mathbf{w}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\by}{\mathbf{y}}
\newcommand{\bz}{\mathbf{z}}

\newcommand{\bF}{\mathbf{F}}
\newcommand{\bV}{\mathbf{V}}
\newcommand{\bX}{\mathbf{X}}

\newcommand{\bxi}{\bm{\xi}}
\newcommand{\bzero}{\bm{0}}

@@

\newcommand{\maxR}{R^{\bm{\oplus}}}
\newcommand{\minR}{R^{\bm{\ominus}}}
\newcommand{\iR}{R^{\bullet}}

@@

We will approximately solve VI \eqref{eq:vi} for certain nonlinear obstacle problems, over two-dimensional computational domains $\Omega \subset \RR^2$, and the Sobolev space $\mathcal{V}=W_0^{1,p}(\Omega)$, using piecewise-linear finite element (FE) spaces over triangular meshes.  The targeted mesh will be the finest level in a hierarchy of meshes constructed by refinement of a coarsest mesh.  In this Section we construct the corresponding multilevel CD.

The mesh hierarchy and FE spaces are constructed in a standard manner \cite{Elmanetal2014}.  Suppose $\Omega$ is polygonal and triangulated by $\mathcal{T}^0$, the \emph{coarsest mesh}, a finite set of non-overlapping triangles with union $\overline{\Omega}$ and characteristic size (e.g.~maximum edge length) $h_0>0$.  Starting with $j=1$, let $\mathcal{T}^j$ be the uniform refinement of $\mathcal{T}^{j-1}$ by edge bisection, thus each $\mathcal{T}^{j-1}$ triangle becomes four similar triangles in $\mathcal{T}^j$ with characteristic size $h_j = 2^{-j} h_0$.  Refinement proceeds to level $J\ge 0$, the \emph{depth} of the hierarchy, giving $\mathcal{T}^J$ as the \emph{finest mesh}.  Let $m_j$ be the number of interior nodes in $\mathcal{T}^j$, the \emph{degrees of freedom} at the $j$th level, and denote $h=h_J=2^{-J} h_0$ and $m=m_J$ for the finest level.  Now let $\mathcal{V}^j$ be the continuous, piecewise-linear FE space $P_1$ \cite{Elmanetal2014} over $\mathcal{T}^j$ with zero boundary values, and observe the following nesting:
\begin{equation}
\mathcal{V}^0 \subset \mathcal{V}^1 \subset \dots \subset \mathcal{V}^J \subset \mathcal{V}.  \label{eq:fe:nestedspaces}
\end{equation}

In the above subspace construction we have chosen triangles and the $P_1$ FE space.  Quadrilaterals and the $Q_1$ FE space will work just as well in what follows.  However, use of higher-order Lagrange elements $P_k$ and $Q_k$ would imply a ``variational crime'' \cite[Chapter 10]{BrennerScott2007} specific to inequality constraints in FE schemes.  Our FE construction of constraint sets requires a property we call \emph{nodal monotonicity}, namely that for all FE functions $w,z \in \mathcal{V}^j$,
\begin{equation}
\bw \ge \bz \quad \implies \quad w \ge z \label{eq:nodalmonotonicity}
\end{equation}
where $\bw,\bz$ denote the vectors of nodal values.  For $k\ge 2$ there are hat functions in $P_k$ and $Q_k$, i.e.~functions with nodal value one at a single node, and otherwise zero nodal values, which take on negative values \cite[Figure 1.7]{Elmanetal2014}, whereas this does not occur for $k\le 1$.  It is easy to show that \eqref{eq:nodalmonotonicity} is equivalent to the non-existence of such sometimes-negative hat functions.

Next we define ``box constraints'' \cite{BensonMunson2006}.  Let $\tilde{\mathcal{V}}^j$ denote the set of functions on the $\mathcal{T}^j$ nodes with values in the extended real line $\tilde{\RR} = [-\infty,+\infty]$.  At the finest-level, suppose that the \emph{lower} and \emph{upper obstacles} $\underline{\gamma}^J, \overline{\gamma}^J \in \tilde{\mathcal{V}}^J$ satisfy
\begin{equation}
\underline{\gamma}^J \le \overline{\gamma}^J, \qquad \underline{\gamma}^J < +\infty, \qquad -\infty < \overline{\gamma}^J, \qquad \text{and} \qquad \underline{\gamma}^J|_{\partial\Omega} \le 0 \le \overline{\gamma}^J|_{\partial\Omega}. \label{eq:fe:boxconstraintrequirements}
\end{equation}
Where $\underline{\gamma}^J=-\infty$ there will no lower constraint, and similarly where $\overline{\gamma}^J=+\infty$ no upper constraint.  Using nodal monotonicity, define
\begin{equation}
\mathcal{K}^J = \left\{v \in \mathcal{V}^J\,:\,\underline{\gamma}^J \le v \le \overline{\gamma}^J\right\}, \label{eq:fineconstraintset}
\end{equation}
the finest-level \emph{constraint set} of \emph{admissible} FE functions, a closed and convex subset of $\mathcal{V}^J$.  As $\mathcal{V}^J$ functions never take values $\pm\infty$, inequality in \eqref{eq:fineconstraintset} is strict where $\underline{\gamma}^J, \overline{\gamma}^J = \pm \infty$.

Let $f^J:\mathcal{K}^J \to (\mathcal{V}^J)'$ be a discretization of $f$ in \eqref{eq:vi}.  Noting that the functional $\ell$ in \eqref{eq:vi} is already defined and continuous over $\mathcal{V}^J$, we denote $\ell^J \in (\mathcal{V}^J)'$ for its FE representation.  We seek the finest-level FE solution $u^J \in \mathcal{K}^J$ of the finite-dimensional, nonlinear VI
\begin{equation}
\ip{f^J(u^J)}{v-u^J} \ge \ip{\ell^J}{v-u^J} \qquad \text{for all } v\in \cK^J. \label{eq:fe:vi}
\end{equation}
As for \eqref{eq:vi}, if $f^J$ is continuous and $p$-coercive on $\mathcal{K}^J$, in the sense that \eqref{eq:pcoercive} holds for all $u,v \in \mathcal{K}^J$, then \eqref{eq:fe:vi} is well-posed.

While \eqref{eq:fe:vi} approximates \eqref{eq:vi}, it may not technically be conforming for \eqref{eq:vi}.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require $\mathcal{K}^J \subset \mathcal{K}$.  The continuum constraint set $\mathcal{K}$ remains abstract in applications where only the fine-level FE obstacles $\underline{\gamma}^J, \overline{\gamma}^J$ are defined.  In fact, even if smooth continuum obstacles exist, admissibility with respect to their FE interpolants might not imply continuum admissiblity.  These caveats specifically apply in geophysical situations where the obstacle is given by meshed elevation data, the continuum limit of which is undefined \cite{Bueler2016}.

The multilevel algorithm in the next section, a solver for VI problem \eqref{eq:fe:vi}, requires transfer operators between meshes.  We use five such operators, three of which are standard, linear maps used for multigrid methods \cite{Trottenbergetal2001}.  Table \ref{tab:transfers} will help the reader keep track.

\begin{table}
\begin{tabular}{llc}
\emph{name}  & \emph{action}  & \emph{linear?} \\ \hline
canonical prolongation        & $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ & \,\checkmark \\
canonical (dual) restriction  & $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ & \,\checkmark \\
(nodal) injection             & $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$ & \,\checkmark \\
maximum restriction           & $\maxR:\tilde{\mathcal{V}}^j\to\tilde{\mathcal{V}}^{j-1}$ & \\
minimum restriction           & $\minR:\tilde{\mathcal{V}}^j\to\tilde{\mathcal{V}}^{j-1}$ & 
\end{tabular}

\medskip
\caption{Transfer operators used in this paper.  The FE vector space $\mathcal{V}^j$ has zero boundary values while $\tilde{\mathcal{V}}^j$ has $\tilde \RR$ values, and arbitrary boundary values.}
\label{tab:transfers}
\end{table}

Canonical prolongation $P:\mathcal{V}^{j-1}\to\mathcal{V}^j$ acts as the identity on coarser-mesh functions in $\mathcal{V}^{j-1}$; it is the linear vector space inclusion $\mathcal{V}^{j-1} \hookrightarrow \mathcal{V}^j$.  Canonical (dual) restriction $R:(\mathcal{V}^j)'\to(\mathcal{V}^{j-1})'$ also acts as an identity in the sense that if $\sigma \in (\mathcal{V}^j)'$ then $(R\sigma)[z] = \sigma[z]$ for all $z \in \mathcal{V}^{j-1} \subset \mathcal{V}^j$.  While these canonical transfers do not lose information, they typically involve a change of FE representation.  For example, the practical formula for canonical restriction, derived by writing coarser-mesh hat functions in terms of finer-mesh ones, is a ``full-weighting'' average \cite{Trottenbergetal2001}, and its transpose is canonical prolongation.

To define injection $\iR:\mathcal{V}^j\to\mathcal{V}^{j-1}$, observe that for any given values at the (interior) nodes of $\mathcal{T}^j$ there is a unique function in $\mathcal{V}^j$ with those values.  Also observe that, by our refinement construction, the nodes of $\mathcal{T}^{j-1}$ are nodes of $\mathcal{T}^j$.  For $z\in\mathcal{V}^j$ we define the injection $\iR z$ as the element of $\mathcal{V}^{j-1}$ which has the same nodal values as $z$ at the nodes of $\mathcal{T}^{j-1}$.  Note that injection is monotonic, that is, if $w \le z$ then $\iR w \le \iR z$.

Finally, the monotone restrictions $\maxR,\minR$ are nonlinear maps which maximize and minimize, respectively, nodal values over the supports of coarser-mesh hat functions.  To make the definitions precise, first denote the \emph{nodes} of triangulation $\mathcal{T}^j$ by $x_q^j$ for $q=0,\dots,\tilde{m}_j-1$, and then the corresponding hat functions \cite{Elmanetal2014} by $\psi_q^j$.  (Boundary nodes are included here.)  The monotone restrictions of $z\in\tilde{\mathcal{V}}^j$ are
\begin{align}
\maxR z &= \sum_{p=1}^{\tilde m_{j-1}} \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}, \label{eq:fe:monotonerestrictions} \\
\minR z &= \sum_{p=1}^{\tilde m_{j-1}} \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}, \notag
\end{align}
with the result in $\tilde{\mathcal{V}}^{j-1}$.  The usual ordering rules for the extended real line $\tilde\RR$ are applied.  Note that if $z\in\tilde{\mathcal{V}}^j$ then $\minR z \le z \le \maxR z$.

@@

To make this precise, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the \emph{defect constraint} \cite{GraeserKornhuber2009}
\begin{equation}
\chi^J = \gamma^J - w^J,  \label{eq:fe:defectconstraintfcn}
\end{equation}
a nonpositive function in $\tilde{\mathcal{V}}^J$.  The motivation here is that a corrected iterate is admissible, $w^J + z^J \ge \gamma^J$, if and only if the perturbation $z^J$ exceeds the defect constraint, $z^J \ge \chi^J$.  Next, for $j=J,J-1,\dots,1$ we apply monotone restriction $\maxR$ to define the \emph{level defect constraints},
\begin{equation}
\chi^{j-1} = \maxR \chi^j. \label{eq:fe:chilevels}
\end{equation}
so that $\chi^j \le \chi^{j-1}$.  A $J=4$ example is shown in Figure \ref{fig:chiphilevels}.

@@

\begin{figure}[ht]
\begin{center}
\input{tikz/nmcdvcycle.tex}
\end{center}
\caption{An NMCD V-cycle computes downward corrections $y_j \in \mathcal{D}^j$, but the upward corrections $z_j$ expand into larger constraint subsets $\mathcal{U}^j$.}
\label{fig:nmcdvcycle}
\end{figure}

@@

\begin{pseudofloat}[H]
\begin{pseudo} \label{ps:nmcd-vcycle}
\pr{nmcd-vcycle}(\ell^J,\gamma^J;w^J)\text{:} \\+
    $\chi^J = \gamma^J - w^J$ \\
    $g^J = w^J$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \maxR \chi^j$ \\
      $\phi^j = \chi^j - P\chi^{j-1}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\phi^j,g^j;y^j)$  \ct{smoothing of \eqref{eq:fe:downvi} in $\mathcal{D}^j$}\\
      $g^{j-1} = \iR(g^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(g^{j-1}) + R \left(\ell^j - f^j(g^j+y^j)\right)$ \\-
    $y^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\chi^0,g^0;y^0)$ \ct{solving of \eqref{eq:fe:downvi} in $\mathcal{D}^0=\mathcal{U}^0$} \\
    $z^0 = y^0$ \\
    for $j=1$ to $j=J$ \\+
      $z^j = P z^{j-1} + y^{j}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\chi^j,g^j;z^j)$  \ct{smoothing of \eqref{eq:fe:upvi} in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\caption{Nonlinear multilevel constraint decomposition V-cycle for the finest-level VI problem \eqref{eq:fe:vi}, as an in-place method on $w^J$.  $f^j$ denotes a discretization of $f$ in problem \eqref{eq:vi}.}
\label{alg:nmcd}
\end{pseudofloat}
